---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Wrangling data to create your first table

Tables can be created as part of a R markdown document. This is because most "table generators" actually converts your code into an output format based on the rules of that particular format. The most common format to work with when you are building your analysis is html. Tables are special since their creation depend more heavily on the output format compared to other elements of R markdown documents. For example, when using the `kable` function from the `knitr` package you have to specify if you want html or latex output. Word output is not possible using the `knitr` package, but other table generators in rmarkdown (such as `flextable`)  *can* produce word output, meaning you can put tables in word documents from your R markdown file. There are several alternatives for generating tables[^tables]. We will start by exploring the built-in function in the `knitr` package.   

[^tables]: See for example the [description of the `gt` package](https://gt.rstudio.com/index.html) that list a bunch of table generators.

The basic work-flow of creating a table in R markdown is to first transform the data into a nice format and then get this underlying data into the table generator. The table generator is written in a code chunk and upon rendering of the R markdown file, the table generator will create, for example, html output.

In this chapter we will also introduce some data wrangling tools since the table we will produce consists of summarized data. The functions we will introduce are found in the pakages `dplyr` and `tidyr`. These packages are loaded as part of the `tidyverse` package.

### Resources

All `tidyverse`packages are [well documented](https://www.tidyverse.org/) and generally well represented in help forums. Google is your friend when looking for help.

The `kable` function is described in a newly developed book, available online called [the R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/kable.html). The package, `kableExtra` comes with excellent [vignettes](https://cran.r-project.org/web/packages/kableExtra/) for both html and pdf outputs. `kableExtra` provides extra functions to customize your basic `knitr` table. 

## Making "Table 1"

The first table in many report in sport and exercise studies is the "Participant characteristics" table. This first table summarizes background information on the participants. We will try to create this table based on data from [@RN2358]. These data can be found in the `exscidata` package. To load the data and other required packages run the following code.

```{r, message=FALSE, warning=FALSE}
library(tidyverse) # for data wrangling
library(knitr) # for table creation
library(kableExtra) # for extra styling of the table
library(exscidata) # the dxadata
```

The end result of this exercise can be found below in Table @\ref(tab:table1-example). This summary table contains the average and standard deviation per group for the variables age, body mass and stature (height) and body fat as a percentage of the body mass. This table is a reproduction of the first part of Table 1 from [@RN2358].


```{r table1-example, echo = FALSE, warning = FALSE, message=FALSE, results = "asis"}

data(dxadata)

dxadata %>%
  select(participant, time, sex, include:weight, fat.whole) %>%
  mutate(fat.whole = ((fat.whole / 1000) / weight) * 100) %>%
  filter(time == "pre") %>%
  group_by(sex, include) %>%
  mutate(n = n()) %>%

  pivot_longer(names_to =  "variable", 
               values_to = "value", 
               cols = age:n) %>%
  group_by(sex, include, variable) %>%
  summarise(m = mean(value), 
            s = sd(value)) %>%
  ungroup() %>%
  mutate(m = signif(m, digits = 3), 
         s = signif(s, digits = 2), 
         ms = if_else(variable == "n", as.character(m), paste0(m, " (", s, ")")), 
         sex_incl = paste(sex, include, sep = "_")) %>%
  dplyr::select(-m, -s, - sex, -include) %>%

  pivot_wider(names_from = sex_incl, 
              values_from = ms) %>%
  select(variable, female_incl, female_excl, male_incl, male_excl) %>%
  mutate(variable = factor(variable, levels = c("n", "age", "weight", "height", "fat.whole"), 
                           labels = c("N", "Age (years)", "Mass (kg)", 
                                      "Stature (cm)", "Body fat (%)"))) %>%
  arrange(variable) %>%
  
  kable(format = "html", 
        col.names = c(" ", "Included", "Excluded", "Included", "Excluded")) %>%
  add_header_above(c(" " = 1, "Female" = 2, "Male" = 2)) %>%
  kable_styling(position = "center")

```

We have to make several operations to re-create this table. First we can select the columns we want to work with further from the data set that also contains a lot of other variables. Let us start by looking at the full data set. Below we use the function `glmipse` from the `dplyr `tibble` package (which is also loaded with `tidyverse`).

```{r}
data("dxadata")

glimpse(dxadata)
```

We can see that we got 80 rows and 59 columns in the dataset. The columns of interest to us are:

- participant
- time
- sex
- include
- age
- height
- weight
- fat.whole

For a full description of the data set, you can type `?dxadata` in your console. The `participant` columns is good to have to keep track of the dataset in the beginning. You might want to have this information to perform "sanity checks" on some results. `time` is needed to remove some observation that are not needed. This pre-training table only sums up information from before the intervention starts. `sex` is a grouping variable together with `include`, Table 1 in [@RN2358] uses Sex and Inclusion in data analysis as grouping for descriptive data. The other variables are used to describe the data sample.

### The pipe operator and select

As mentioned above, we will start by selecting the variables we want to work further with. Using the `select` function from dplyr we can select columns that we need. In the code below I will use select as part of a "pipe". Think of the pipe as doing operations in sequel. Each time you use the pipe operator (`%>%`) you say "then do". The code below translates to: 

- Take the dataset `dxadata`, *then do*
- `select` the following variables, *then do*
- print

`print`, is a function that outputs the results of the operations. In each new function of a pipe, the data that we take with us from the above line ends up as the first argument. A representation of this behavior can be expressed as:

`DATA %>%
  FUNCTION(DATA, ARGUMENTS) %>%
  FUNCTION(DATA, ARGUMENTS) %>%
  FUNCTION(DATA)`
  
We do not need to type the data part, instead the pipe operator (`%>%`) gathers the data from each step and puts it in the subsequent function.

Copy the code below to your own script or Rmarkdown document and run it. If you use a Rmarkdown document (.Rmd), you might want to set "Chunk output in console" in the settings menu. In my experience, this makes developing code a bit faster. 

```{r, eval = FALSE}

dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  print() # print the output

```

Notice that I have added short comments after each line to make it clear what I want to accomplish. We will build further on the above code, and this is a common work-flow. Using pipes, it is easy to extend the code by adding lines doing certain operations, one at the time. Notice also that the select function uses a list of variable names with `include:weight` being short for "take all variables *from* include *to* weight.

### Filter observations

The next step will be to filter observations. We need to remove the observations that comes from the post-intervention tests. The `time` variable contains to values `pre` and `post` to remove post-values we need to tell R to remove these all observations (rows) containing `post`. We will use the `filter` function from dplyr. This will be our first experience with logical operators. Let's try out to alternatives, copy the code to your console to see the results.

```{r, eval = FALSE}
## Alternative 1: ##
dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter away all observation with "post"
  filter(time != "post") %>%
  
  print() # print the output


## Alternative 2: ##
dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  print() # print the output



```

The above code should give the same output. The operator `!=` says "not equal to", the operator `==` says equal to. Notice that R uses two equal signs to say equal to. A single equal sign is used as an assignment operator in R.

### Create or change variables

The next problem for us is that we need to manipulate or combine information from two variables in order to calculate body fat percentage. The formula that we will use is simply expressing body fat as a percentage of the body weight.

$$Body~fat~(\%) = \frac{Body~fat~(g)/1000}{Body~weight~(kg)} \times 100$$
By using the mutate function we can add or manipulate existing variables in a pipe. Mutate takes as arguments a list of new variables:

```{r, eval = FALSE}


dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  print() # print the output


```

In the code above, we "replace" the variable `fat.whole` with the re-calculated variable.

### Grouped operations and summary statistics

In a pipe, we can group the data set giving us opportunities to calculate summary statistics over one or several grouping variables. In Table 1 in [@RN2358], `include` and `sex` are the two grouping variables. Using the `group_by()` function from `dplyr` sets a grouping of the data frame. If we use functions that summarizes data, such summarizes will be per group. In Table 1 in [@RN2358] the number of participants in each group are specified. We can use the function `n()` to calculate the number of observations per group in a `mutate` call.

```{r, eval = FALSE}
dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  
  # Group the data frame and add a variable specifying the number of observations per group
  group_by(include, sex) %>%
  mutate(n = n()) %>%
  print() # print the output


```

The new variable `n` now contains the number of observations in each group. For now we can regard this as a new variable. Each participant belongs to a specified group, and this specific group has `n` number of members.

We can now go further and use the `summarise` function. Instead of adding variables to the existing data set, summarize reduces the data set using some summarizing function, such as `mean()` or `sd()`. These estimates are what we are looking for in our data set. Example of other summarizing functions for descriptive data are `min()` and `max()` for the minimum and maximum.

We can use the `summarise()` function to calculate the mean and standard deviation for the weight variable.

```{r, eval = FALSE}

dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  
  # Group the data frame and add a variable specifying the number of observations per group
  group_by(include, sex) %>%
  mutate(n = n()) %>%
  # Summarise weight
  summarise(weight.m = mean(weight), 
            weight.s = sd(weight)) %>%
  
  print() # print the output
```

Try out the code in your own script. The above example gives us what we want, however, it means that we need to type a lot. Instead of needing to make a summary for each variable, we can combine the variables in a long format. To get to the long format we will use the `pivot_longer()` function. This function gathers several variables into two columns, one with the variables names as values and a second column with each value from the original variables. In our case we want to gather the variables `age`, `height`, `weight`, `fat.whole` and `n`. I will call the new variables that we create `variable` and `value`.

```{r, eval = FALSE}

dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  
  # Group the data frame and add a variable specifying the number of observations per group
  group_by(include, sex) %>%
  mutate(n = n()) %>%
  # Collect all variables for convenient summarizing
  pivot_longer(names_to = "variable", 
               values_to = "value", 
               cols = age:n) %>%
  print()



```

The `cols = age:n` part of pivot_longer specifies what columns to gather. The dataset is still grouped by `include` and `sex`. We may now proceed by summarizing over these groups, however, we need to add another group to specify that we want differnt numbers per variable. To do this we re-specify the grouping. After this we add the `summarize` function. 

```{r, eval = FALSE}

dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  
  # Group the data frame and add a variable specifying the number of observations per group
  group_by(include, sex) %>%
  mutate(n = n()) %>%
  # Collect all variables for convenient summarizing
  pivot_longer(names_to = "variable", 
               values_to = "value", 
               cols = age:n) %>%
  # Create a new grouping, adding variable
  group_by(include, sex, variable) %>%
  # Summarize in two new variables m for mean and s for SD
  summarise(m = mean(value), 
            s = sd(value)) %>%
  print()

```

If you run the above code you will notice that the the standard deviation of each variable is larger than zero except for `n` which has no variability. This is because we created it per group and simply calculated it as the sum of observations.

Take a look at [Table 1](https://physoc.onlinelibrary.wiley.com/doi/full/10.1113/JP278455) in [@RN2358]. The format of the descriptive statistics are mean (SD), this is a preferred way of reporting these estimates. In order to achieve this we need to "manually" convert the numbers. In the example below, I will start by making a new variable by simply pasting the numbers together. I will also add the brackets.

```{r, eval = FALSE}

dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  
  # Group the data frame and add a variable specifying the number of observations per group
  group_by(include, sex) %>%
  mutate(n = n()) %>%
  # Collect all variables for convenient summarizing
  pivot_longer(names_to = "variable", 
               values_to = "value", 
               cols = age:n) %>%
  # Create a new grouping, adding variable
  group_by(include, sex, variable) %>%
  # Summarize in two new variables m for mean and s for SD
  summarise(m = mean(value), 
            s = sd(value)) %>%
  # Add descriptive statistics together for nice formatting
  mutate(ms = paste0(m, " (", s, ")"))
  print()



```

In `mutate(ms = paste0(m, " (", s, ")"))`, the `paste0` function simply paste components together to form a string of text. First, the *vector* of means are being used, then we add a parenthesis, followed by the SD and finally a parenthesis.

If you run the above code you will notice that you end up with numbers looking like this:

`167.666666666667 (6.86851298231541)`

This is neither good or good looking. We have to take care of the decimal places. There are a number of ways to do this but in this case the function `signif` seems to make the situation better. `signif` rounds to significant digits. This means that we will get different rounding dependning on the "size" of the value. I find `signif(m, 3)` to be agood starting point.

```{r, eval = FALSE}


dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  
  # Group the data frame and add a variable specifying the number of observations per group
  group_by(include, sex) %>%
  mutate(n = n()) %>%
  # Collect all variables for convenient summarizing
  pivot_longer(names_to = "variable", 
               values_to = "value", 
               cols = age:n) %>%
  # Create a new grouping, adding variable
  group_by(include, sex, variable) %>%
  # Summarize in two new variables m for mean and s for SD
  summarise(m = mean(value), 
            s = sd(value)) %>%
  # Add descriptive statistics together for nice formatting
  mutate(ms = paste0(signif(m, 3), # Use signif to round to significant numbers
                     " (",
                     signif(s, 3), 
                     ")")) %>%
  print()



```

Things are starting to look good. Run the code, what do you think. A problem with the above is that we do not want any dispersion/variability in the `n` variable. So if the variable is `n`we do not want that kind of formatting. It is time to add a conditional statment. In `dplyr` there are easy-to-use if/else functions. The function `if_else` sets a condition, if this is met then we can decide what to do, and likewise decide what to do if it is not met.

This looks something like this inside a `dplyr` pipe:

```{r, eval = FALSE}

... %>%
  if_else(IF_THIS_IS_TRUE, THE_DO_THIS, OTHERWISE_DO_THIS) %>%
  print()

```

If `variable` is `n`, then we only want to display `m` otherwise we want the full code as described above: `paste0(signif(m, 3), " (", signif(s, 3), ")")`. We add this to the code:


```{r, eval = FALSE}


dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  
  # Group the data frame and add a variable specifying the number of observations per group
  group_by(include, sex) %>%
  mutate(n = n()) %>%
  # Collect all variables for convenient summarizing
  pivot_longer(names_to = "variable", 
               values_to = "value", 
               cols = age:n) %>%
  # Create a new grouping, adding variable
  group_by(include, sex, variable) %>%
  # Summarize in two new variables m for mean and s for SD
  summarise(m = mean(value), 
            s = sd(value)) %>%
  # Add descriptive statistics together for nice formatting
  mutate(ms = if_else(variable == "n", # If the variable is n
                      as.character(m), # the only display the mean, otherwise:
                        paste0(signif(m, 3), # Use signif to round to significant numbers
                        " (",
                        signif(s, 3), 
                        ")"))) %>%
  print()



```

The `as.character` part is needed because the output of `if_else` must be the same regardless of what the outcome of the test is.

We are getting close to something!

The next step is to remove variables that we do not longer need. The `select` function will help us with that. we can remove `m`and `s` by `select(-m, -s)`, the minus sign tells are to remove them from the list of variables in the dataset. We can then combine the grouping variables into a `include_sex` variable. Similarly to what we did above, we can simply paste them together. Now we will use the `paste` (function instead of `paste0`). In `paste` we specify a separator, maybe `_` is a nice alternative. Selecting away the individual variables from the new combined one leaves us with this code and data set.

```{r, eval = FALSE}


dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  
  # Group the data frame and add a variable specifying the number of observations per group
  group_by(include, sex) %>%
  mutate(n = n()) %>%
  # Collect all variables for convenient summarizing
  pivot_longer(names_to = "variable", 
               values_to = "value", 
               cols = age:n) %>%
  # Create a new grouping, adding variable
  group_by(include, sex, variable) %>%
  # Summarize in two new variables m for mean and s for SD
  summarise(m = mean(value), 
            s = sd(value)) %>%
  # Add descriptive statistics together for nice formatting
  mutate(ms = if_else(variable == "n", # If the variable is n
                      as.character(m), # the only display the mean, otherwise:
                        paste0(signif(m, 3), # Use signif to round to significant numbers
                        " (",
                        signif(s, 3), 
                        ")")), 
         # Doing a new grouping variable
         include_sex = paste(include, sex, sep = "_")) %>%
  # removing unnecessary variables after ungrouping
  ungroup() %>%
  select(-sex, -include, -m, -s)
  print()

```

If `ungroup` is not used, we cannot select away variables since they are used to group the dataset. We will now perform the last operations befor we can make it a table. To make it formatted as in Table 1 in [@RN2358], we can make the present dataset wider. Each group has its own variable in addition to the variable name column. We will use the opposite function to `pivoy_longer`, namely `pivot_wider`[^pivot]. `pivot_wider` takes a variable or "key" column and a "values" column and divide the values based on the "key".

[^pivot]: All this talk about pivot, take a break and [watch this clip](https://www.youtube.com/watch?v=L_PWbnHABsM) from the hit series "Friends", its about "pivot"!

```{r, eval = FALSE}

dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  
  # Group the data frame and add a variable specifying the number of observations per group
  group_by(include, sex) %>%
  mutate(n = n()) %>%
  # Collect all variables for convenient summarizing
  pivot_longer(names_to = "variable", 
               values_to = "value", 
               cols = age:n) %>%
  # Create a new grouping, adding variable
  group_by(include, sex, variable) %>%
  # Summarize in two new variables m for mean and s for SD
  summarise(m = mean(value), 
            s = sd(value)) %>%
  # Add descriptive statistics together for nice formatting
  mutate(ms = if_else(variable == "n", # If the variable is n
                      as.character(m), # the only display the mean, otherwise:
                        paste0(signif(m, 3), # Use signif to round to significant numbers
                        " (",
                        signif(s, 3), 
                        ")")), 
         # Doing a new grouping variable
         include_sex = paste(include, sex, sep = "_")) %>%
  # removing unnecessary variables after ungrouping
  ungroup() %>%
  select(-sex, -include, -m, -s) %>%
  # pivot wider to match the desired data
  pivot_wider(names_from = include_sex, 
              values_from = ms) %>%
  print()

```

A final step is to format the `variable` variable(!). The easiest is to make it a factor variable with specified levels and names. After we have added this information, we can use `arrange` to sort the dataset. Using `select` will help you sort the columns to match wath we want.

```{r, eval = FALSE}

dxadata %>% # take the dxadata data set
  select(participant, time, sex, include:weight, fat.whole) %>% 
  # select participant, time, sex, include to height and fat.whole
  
  # Filter to keep all observations with pre
  filter(time == "pre") %>%
  
  # Calculate body fat
  # fat.whole in grams, needs to be divided by 1000 to express as kg
  # Multiply by 100 to get percentage
  mutate(fat.whole = ((fat.whole/1000) / weight) * 100) %>%
  
  # Group the data frame and add a variable specifying the number of observations per group
  group_by(include, sex) %>%
  mutate(n = n()) %>%
  # Collect all variables for convenient summarizing
  pivot_longer(names_to = "variable", 
               values_to = "value", 
               cols = age:n) %>%
  # Create a new grouping, adding variable
  group_by(include, sex, variable) %>%
  # Summarize in two new variables m for mean and s for SD
  summarise(m = mean(value), 
            s = sd(value)) %>%
  # Add descriptive statistics together for nice formatting
  mutate(ms = if_else(variable == "n", # If the variable is n
                      as.character(m), # the only display the mean, otherwise:
                        paste0(signif(m, 3), # Use signif to round to significant numbers
                        " (",
                        signif(s, 3), 
                        ")")), 
         # Doing a new grouping variable
         include_sex = paste(include, sex, sep = "_")) %>%
  # removing unnecessary variables after ungrouping
  ungroup() %>%
  select(-sex, -include, -m, -s) %>%
  # pivot wider to match the desired data
  pivot_wider(names_from = include_sex, 
              values_from = ms) %>%
    mutate(variable = factor(variable, levels = c("n", "age", "weight", "height", "fat.whole"), 
                           labels = c("N", "Age (years)", "Mass (kg)", 
                                      "Stature (cm)", "Body fat (%)"))) %>%
  arrange(variable) %>%
  print()

```

### Starting the table generator - The `kable` function

The next step is to push everything into the table generator. `kable` from the `knitr` package creates the framework for the table we are going to create. Next we will use functions from the `kableExtra` package to customize the table. Use the help pages for `kable` to see what can be specified in the function. We will make use of only a few arguments. `kable` can be part of a pipe, the same as the following `kableExtra` functions. 



### Prerequisistes

Load packages with the following commands:

```{r load_packages, eval = FALSE}
library(tidyverse) # for data wrangling and file download
library(knitr) # for the kable function
library(kableExtra) # for extended capabilities
```

Download the example file with the following command:

```{r download_data, eval=FALSE, message=TRUE, warning=TRUE}



```

## First steps, decide what data to use and summarise

Before we can start to format the table, we need to decide what data to put in it. The goal here is to create a summary table of participant characteristics. We can group the participants based on inclusion/exclusion from the study and sex.

Above, we stored the data in an object called `dxa_data`. Use the below code to see what it contains.

```{r data exploration1, eval = FALSE}
dxa_data %>%
        print()
```

This is data from iDXA measurements that also contains weight, height, age, sex and if participants were included in the data analysis or not (`include == "incl"` or `include == "excl"`). First we want to use lean body mass as a percentage of the whole body mass as a measure to describe the participants. The below code does the calculation. We also select columns of interest for further data preparations using the `select()` function. Store the data in a new object, so that the raw data can be used later. 

```{r calculate_LBM, eval = FALSE}
dxa <- dxa_data %>%
        rowwise() %>%
        mutate(LBM = (lean.whole / (fat.whole + BMD.whole + lean.whole)) * 100) %>%
        select(subject, age, height, weight, LBM, sex, include) %>%
        print()
```

Using `dplyr::group_by()` and `dplyr::summarise()` we can now create a summary table.

> When writing `dplyr::group_by()` it means that we specify the function `group_by()` from the package `dplyr`. The double colon is used to connect the package to the function, we use it here to specify what package we are using.

```{r data summary, eval = FALSE}
dxa %>%
        ungroup() %>%
        group_by(sex, include) %>%
        summarise(m.age = mean(age), 
                  s.age = sd(age), 
                  m.height = mean(height), 
                  s.height = sd(height), 
                  m.weight = mean(weight), 
                  s.weight = sd(weight), 
                  m.lbm = mean(LBM), 
                  s.lbm =sd(LBM)) %>%
        print()

```

This was somewhat inefficient. We can add a line and calculate per variable.

```{r data summary2, eval = FALSE}
summary_table <- dxa %>%
        ungroup() %>%
        pivot_longer(cols = age:LBM, names_to = "variable", values_to = "value") %>%
        group_by(sex, include, variable) %>%
        summarise(m = mean(value), 
                  s = sd(value)) %>%
        print()
```

This was more efficient! But it's not formatted to be a table just yet. Of course there are many ways to create this table. But first, we can format the numbers to make nice presentations. For simplicity, lets say that all numbers will be printed with a single digit. We could use `round(x, 1)` but this will in some cases produce a number without trailing zero. Instead we will use `sprintf("%.1f", x)` where x is the number of interest. We will also combine mean and standard deviation to a single cell using the `paste0()` function. `paste0()` combines numbers and text into a character string, this means they will lose some information but that is OK.

```{r data summary3, eval = FALSE}
summary_table <- dxa %>%
        ungroup() %>%
        pivot_longer(cols = age:LBM, names_to = "variable", values_to = "value") %>%
        group_by(sex, include, variable) %>%
        summarise(m = mean(value), 
                  s = sd(value)) %>%
        ungroup() %>%
        mutate(summary = paste0(sprintf("%.1f", m),
                                " (",
                                sprintf("%.1f", s), 
                                ")")) %>%
        select(sex, include, variable, summary) %>%
        pivot_wider(id_cols = variable, names_from = c(sex, include), values_from = summary ) %>%
        print()
```

The summary table is ready for making a table. 

In R Markdown chunk settings, we need to set the output format to `results = "asis"`. This will print the table correctly. To create the table, all we need to do is to is to pipe it to `kable()`. In the `kable` function we can set `caption` to the represent the title of the table. `col.names` lets you specify column names without changing variable names, this is handy when you want the same name to repeat many times.     


````
`r ''````{r my_table, results="asis"}
summary_table %>%
        kable(format = "html", col.names = c("Variable", 
                                                "Female excluded",
                                                "Female included", 
                                                "Male excluded", 
                                                "Male included"), 
                                        caption = "Participant characteristics")
```
````

We might want to extend the table and make it a bit nicer, this is where `kableExtra` comes in. Lets say that we want to include a row above Female/Male and only print this information once. After the `kable()` command, we use commands from the `kableExtra` package.  



```{r kableExtra,  eval = FALSE}
summary_table %>%
        kable(format = "html", col.names = c(" ", 
                                                "Excluded",
                                                "Included", 
                                                "Excluded", 
                                                "Included"), 
                                        caption = "Participant characteristics") %>%
                                        add_header_above(c(" " = 1, "Female" = 2, "Male" = 2))
```

There is still some work to do here, variable names should be with a capital letter, this is most easily done in data preparation stages by changing the variables factor. 

```{r mytable_cont1, eval = FALSE}
dxa %>%
        ungroup() %>%
        pivot_longer(cols = age:LBM, names_to = "variable", values_to = "value") %>%
        group_by(sex, include, variable) %>%
        summarise(m = mean(value), 
                  s = sd(value)) %>%
        ungroup() %>%
        mutate(summary = paste0(sprintf("%.1f", m),
                                " (",
                                sprintf("%.1f", s), 
                                ")")) %>%
        select(sex, include, variable, summary) %>%
        pivot_wider(id_cols = variable, names_from = c(sex, include), values_from = summary ) %>%
        # sort the rows -- create a factor level
        mutate(variable = factor(variable, levels = c("age", "height", "weight", "LBM"))) %>%

        mutate(Variable = c("Age (yrs)", "Stature (cm)", "Body mass (kg)", "Lean body mass (%)")) %>%
        
        arrange(variable) %>%
        
        select(Variable, female_excl:male_incl) %>%
        kable(format = "html", col.names = c(" ", 
                                                "Excluded",
                                                "Included", 
                                                "Excluded", 
                                                "Included"), 
                                        caption = "Table 1. Participant characteristics") %>%
                                        add_header_above(c(" " = 1, "Female" = 2, "Male" = 2)) 
```

kableExtra lets us add  footnotes using the `footnote` function. An example of the whole pipe could be:

```{r mytable_cont2, eval = FALSE}
dxa %>%
        ungroup() %>%
        pivot_longer(cols = age:LBM, names_to = "variable", values_to = "value") %>%
        group_by(sex, include, variable) %>%
        summarise(m = mean(value), 
                  s = sd(value)) %>%
        ungroup() %>%
        mutate(summary = paste0(sprintf("%.1f", m),
                                " (",
                                sprintf("%.1f", s), 
                                ")")) %>%
        select(sex, include, variable, summary) %>%
        pivot_wider(id_cols = variable, names_from = c(sex, include), values_from = summary ) %>%
        # sort the rows -- create a factor level
        mutate(variable = factor(variable, levels = c("age", "height", "weight", "LBM"))) %>%
        arrange(variable) %>%
        mutate(Variable = c("Age (yrs)", "Stature (cm)", "Body mass (kg)", "Lean body mass (%)")) %>%
        select(Variable, female_excl:male_incl) %>%
        kable(format = "html", col.names = c(" ", 
                                                "Excluded",
                                                "Included", 
                                                "Excluded", 
                                                "Included"), 
                                        caption = "Table 1. Participant characteristics") %>%
                                        add_header_above(c(" " = 1, "Female" = 2, "Male" = 2)) %>%
                                        footnote(general = "Values are Mean (SD)")
```

## Final thoughts on `knitr::kable` and `kableExtra`

The [vignette describining the kableExtra package](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) is highly recommended! If you want to extend to pdf-tables (which are really nice!). Use the $\LaTeX$ vignette found [here](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_pdf.pdf). When creating pdf/$\LaTeX$ tables/reports some basic understandning of $\LaTeX$ is good. 


## Flextable

Flextable comes with the option of specifying tables directly in word outputs. This does not need any further formatting compared to HTML formatting. 

Lets use the `summary_table` to specify the flextable. Flextable also works as part of a pipe. The basic function is `flextable`.

```{r, eval = FALSE} 

library(flextable)

summary_table %>%
   flextable()

```

We can now add changes to the underlying table and add new header labels to produce a table similar to the one above.

```{r, eval = FALSE} 

library(flextable)

summary_table %>%
        mutate(variable = factor(variable, levels = c("age", "height", "weight", "LBM"), 
                                          labels = c("Age (yrs)", 
                                          "Stature (cm)", 
                                          "Body mass (kg)",
                                          "Lean body mass (%)"))) %>%
   flextable()  %>%
   set_header_labels(variable = "",
                    female_excl = "Exclude", 
                     female_incl = "Include",
                     male_excl = "Exclude", 
                     male_incl = "Include")

```

Note that flextable does not need `output = "asis"` to work in markdown files. Similar to kable, flextable can give a preview by just running the code, meaning you do not have to knit the whole document to see what changes you have done.

Lets finnish up by adding a header and a footnote.

```{r, eval = FALSE} 

library(flextable)

summary_table %>%
        mutate(variable = factor(variable, levels = c("age", "height", "weight", "LBM"), 
                                          labels = c("Age (yrs)", 
                                          "Stature (cm)", 
                                          "Body mass (kg)",
                                          "Lean body mass (%)"))) %>%
   flextable()  %>%
   set_header_labels(variable = "",
                    female_excl = "Exclude", 
                     female_incl = "Include",
                     male_excl = "Exclude", 
                     male_incl = "Include") %>%
      # Adds a header specified for all columns of the table
    add_header_row(values = c("", "Female", "Female", "Male", "Male")) %>%
    
    # Merge the new cells with the same data
    # part = "header" meands that we want to add a row in the "header" part of the table.
    # i = 1 means row 1
    # j = 2:3 means column 2 to 3
    merge_at(part = "header", i = 1, j = 2:3) %>%
    merge_at(part = "header", i = 1, j = 4:5) %>%
    # Add footnote
    add_footer_row(values = "Values are mean and (SD)", colwidths = 5) %>%
    # Make the columns widths match the content
    autofit()
    
    
    

```

Nice! This table should print if the code chunk is included in a R Markdown file and the output is choosen to be "Knit to word". See [flextable documentation](https://davidgohel.github.io/flextable/) for more customizations.



