---
title: "Associations and causation"
editor_options: 
  chunk_output_type: console
---

We have previously been introduced to a general purpose algorithm for describing associations among variables, the regression model. From a univariate regression model we can also get the correlation coefficient which tells us to which degree two variables are associated on a unit-less scale between -1 and +1.

We can further extend our definition of association by including other types of data variables in our discussion. A discrete variable, such as sex, may be associated with body weight. We could assess this by examining the average body weight among males and females. Sex could also be associated with the prevalence of Anterior Cruciate Ligament (ACL) Injuries. We could assess this by looking at the occurrence of injuries (injured/not injured) among males and females. This would be an investigation of association between two discrete variables. Similarly, discrete variables with more than two levels can be analyzed for association with other discrete or continuous variables.

However, we must be aware, as *association does not imply causation*. Assuming that a variable is causally associated with another variable may guide us in making decision. For example, decisions on how we should organize our society, or our everyday life. If assuming causality from association USA should decrees funding of science, space and technology as this is highly, and positively correlated with suicides by hanging, strangling and suffocation (@fig-spurious-correlation)[^corr-sciencefunding]. Decreasing science funding would lead to less suicides by hanging, if these phenomena were *causally* associated. They are likely not.  

[^corr-sciencefunding]: See [here](https://www.tylervigen.com/spurious-correlations)

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-spurious-correlation
#| fig-cap: "US spending on science, space and technology is positively related to the number of suicied by hanging, strangulation and suffocation."

library(tidyverse)

data.frame(suicides = c(5427, 5688, 6198, 6462, 6635, 7336, 7248, 
                        7491, 8161, 8578, 9000), 
           spending = c(18.079, 18.594, 19.753, 20.734, 20.831, 23.029, 
                        23.597, 23.584, 25.525, 27.731, 29.449), 
           year = seq(from = 1999, to = 2009)) %>%
  ggplot(aes(spending, suicides)) + geom_point() +
  
  labs(x = "US spending on science, space and technology (billion $)", 
      y = "Suicides by hanging,\nstrangulation and suffocation") + 
  theme_minimal()


```

## Drawing a scientific model using Directed Acyclic Graphs

Using Directed Acyclic Graphs (DAGs) we can describe causal assumptions of a system that we are investigating. DAGs can be used to describe nonparametric causal relationships. The graph consists of nodes and arrows (or edges). Each node encodes a variable and arrows shows the direction of the causal effect. @fig-dag-simple shows three possibilities 


```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-dag-simple
#| fig-cap: "Chain, fork and inverted fork"

library(ggdag)

dagify(y ~ x, 
       x ~ z, 
       m ~ k, 
       t ~ k, 
       v ~ s, 
       s ~ d) %>%
  ggdag() + 
  theme_void()




```

From 







- Chains 
- Forks
- Inverted forks

- Confounding
- Collider bias
- Overcontrol bias 

- Descendants
- Ancestors

- Back-door paths

- Post-treatment variables

## Causality and randomized experiments



```{r}

a <- rnorm(100, 0, 1)

c <- rnorm(100, 0, 1)

b <- a*0.5 + c*0.5 + rnorm(100, 0, 1)


summary(lm(a ~ c + b))


plot(a,c)
plot(b,c)


```






