---
output: html_document
bibliography: resources/bib-final.bib
editor_options: 
  chunk_output_type: console
---

# Assignment 5: Analyzing repeated measures experiments {#assignment5}

In this assignment you will analyze and report on trial investigating the effect of resistance training volume on lean mass and muscle strength. The data are part of the `exscidata` package and can be accessed as `data("strengthvolume")` and `data("dxadata")`. Read the instructions carefully!

## Organizing the report

Your report should consist of the sections Introduction, Methods, Results and Discussion. Each part of the report should be written as a reproducible document and a link or reference to the repository containing the source document(s) should be included in the report. Below follows detailed descriptions and requirements for each section.

### Introduction

This section should consist of a description of the field, resistance-training volume and muscle strength and mass. Use at least five to ten references to introduce your audience and explain why you are doing the analysis/study. A tip is to use the QALMRI method, introduced in Assignment 4 (See chapter \@ref(assignment4)) to structure the reading of background information. It is up to you how you motivate the study and how you phrase the purpose of the study. It could be a hypothesis based on previous studies, it could also be question to fill a knowledge gap that you have identified in your literature review.

Structure the introduction in paragraphs. A first paragraph could contain a general introduction to the field, why is it of interest to investigate resistance-training? A second paragraph could specifically describe the specific field of resistance-training volume, why is important to know more about how we are likely to respond to different training volumes. The second paragraph should incorporate definitions important for your report, e.g., training volume, muscle mass and strength. Try to incorporate these definition as a fluid part of the text.

A third (or last) paragraph of the introduction should contain a statement regarding the purpose of the study. The purpose could be descriptive, hypothesis-driven or guided by a question. Although it could be considered a bit backward, you should explore the data sets before you select your question/hypothesis/purpose for it to be possible to answer.

### Methods

The method should give a thorough overview of the study and specific details regarding data collection. You can read about the details of this specific study in [@RN2358]. Use your own words to describe the study based on this description. A nice way to structure the methods section is to include subheadings:

-   Participants and study overview: Describe the participants and give an overview of all tests/measurements. Participants should be described in the first table of the report (Table 1). The overview of the tests/measurements should be done without double presentation as details should be presented in subsequent sections.

-   Specific descriptions (e.g. strength tests): Describe in detail how tests/measurements that you mentioned in the overview where conducted.

-   Data analysis and statistics: Describe how you treated the data prior to statistical tests or procedures and what tests/procedures were used to draw inference (or more generally, to answer your purpose). Describe how you present data (e.g. descriptive data with mean (SD), inference with confidence intervals etc.).

### Results

Describe the results of your analysis. This description should make use of table and figures as well as a text that guides and structures the content to the reader. Think about it this way, the text should describe when and how to read the figures and tables. This means that all aspects of the results should be covered in the text. The figures and tables should also be "self explanatory", this means that you have to include descriptive figure captions and descriptions of tables (see below for tips).

As the main purpose of the analysis should concern the effect of training volume on muscle mass and strength, it is natural that the comparison of training outcomes between volume conditions is the main analysis in the results. You may also have questions regarding the relationship between muscle strength and mass gains, if there are differences between men and women etc. Selection of statistical/analysis techniques should reflect the study question/purpose.

### Discussion

Structure the discussion with a first paragraph describing the main results of the analysis, this could be the answer to your question or a statement regarding the study hypothesis. In the following paragraphs discuss all results that you have presented in the light of previous studies. It is your job to give the reader plausible interpretations and explanations of your results. This is how single scientific results are incorporated in our collective understanding. These interpretations can later be challenged, however if you give the reader good arguments and clear descriptions, your insights will be valuable to collective reasoning even if they turn out to be wrong in light of new data.

End the discussion with a summary or conclusion. Some journals request that you state your conclusions under a specific heading in the end of the report/article.

## Organizing the data analysis

### Data preparation

The data is already structured in the `exscidata` package. To access the data, use the following code:

```{r, eval = TRUE, echo = TRUE}

library(exscidata)
data("dxadata"); data("strengthvolume")


```

To get and overview of the variables in each data set use `?strengthvolume` and `?dxadata`. In the `dxadata` the variables of interest are organized in a more convenient way using the code below:

```{r, message = FALSE, warning = FALSE, echo = TRUE}

library(tidyverse)

dxadata %>%
  select(participant:include, lean.left_leg, lean.right_leg) %>%
  pivot_longer(names_to = "leg", 
               values_to = "lean.mass", 
               cols = lean.left_leg:lean.right_leg) %>%
  mutate(leg = if_else(leg == "lean.left_leg", "L", "R"), 
         sets = if_else(multiple == leg, "multiple", "single")) %>%
  select(participant, time, sex, include, sets, leg, lean.mass) %>%
  print()

```

You are free to use any other part of the data set in the assignment.

In the `strengthvolume` data set, strength measurement are made in several different exercises. In [@RN2358], one analysis was done on a combined score. This score was calculated as the scaled average. To obtain this combined score, use the code below:

```{r, message = FALSE, warning = FALSE, echo = TRUE}

strengthvolume %>%
  group_by(exercise) %>%
  mutate(scaled.load = load / max(load, na.rm = TRUE)) %>%
  group_by(participant, time, sex, include, sets) %>%
  summarise(combined.load = mean(scaled.load, na.rm = TRUE)) %>%
  ungroup() %>%
  print()

```

The combined load calculated above is a measure of strength without unit (it is not e.g. kg). It can be used to calculate percentage or raw change. You may however choose to only analyze one type of strength measurement.

In the original analysis of this data, the maximum from the first two measurements were used as the baseline. It is up to you how you treat the data and what part of the data you use for your analysis. Most included participants have a measure of strength at pre/session1 and post. 18 participants have a complete data set with strength measured at weeks 2, 5, and 9 in addition to pre/post.

Both data sets require more exploration for you to get fully familiar with them. Make sure you make these explorations outside the final report.

### Statistical models

It is up to you what model you choose to analyze your data. You have to explain your choice of model, but you may keep it simple. If you use an ANCOVA model, remember that participants did both training conditions, this means that you will have to include a random effect giving information of participant ID to the model. Your model will be a mixed-ANCOVA. If you use `lme4`, the model formulation could look like this:

`change ~ baseline + sex + sets + (1|participant)`

If you use a mixed model approach you may use raw scores or e.g. log-transformed raw-scores. Your model could look like this:

`lean.mass ~ time * sets + (1|participant)`

In the ANCOVA model, sex is important to account for differences in the baseline strength. Sex could explain some variation in the mixed-model but is unlikely to change the results regarding the `time:sets` interaction.

You may extend or change these model formulations as long as these changes are explained and motivated by your question/purpose/hypothesis. Additional statistical tests/models such as correlation or regression models should be described and motivated by questions/purposes in your report.

### Tools for writing the report

By now, you should be familiar with writing in Rmarkdown. RStudio provides a nice interface in the "visual markdown editor" that also provides integration with citation managers and bibliographies. We have introduced `knitr` and `kableExtra` as two packages that you can use to create tables. You should also be familiar with making figures using `ggplot2`. In previous chapters we have mentioned `cowplot` as a way to combine several `ggplot2` objects in one plot.

To fit models requiring random effects `lme4` is the package you have been introduced to. Estimates (group/condition averages) from these models can be obtained using the `emmeans` package.

To format data you have been introduced to `dplyr` and `tidyr`, both packages are part of the `tidyverse` package.

## Handing in the report

As mentioned above, the report should have be reproducible as it should be built from a source `Rmd` file. In the report, include a statement regarding where to find the source document. This may be a link to your github repository. The report should be handed in for comments and then included in the portfolio exam.
