---
output: html_document
bibliography: resources/bib-final.bib
editor_options: 
  chunk_output_type: console
---

# Assignment 1: Reliability and tools for reproducible data science {#assignment1}

The purpose of this assignment is to present estimates of reliability of measures collected in the physiology lab. A second purpose is to use tools for reproducible data science. The report that you are expected to hand in therefore has some strict requirements in its format (see below). The assignment is a group assignment and at least three students are expected to contribute to each report.

## Elements of the report

The report should describe one test that you have performed in the physiology-lab. Select the test that is most interesting to you. The test should be described with a detailed protocol, including preparations of the participant (that is being tested), standardization of the test, and post-test data preparation. Post-test data preparation refferes to steps needed to get data from e.g. equipment used during the test. This section should take into account and reference [@RN1739; @RN2511]

The next section should contain descriptive data from the test. This could be measures of central tendency and variability in the measures you have collected. If possible, try to find similar estimates in the scientific literature.

Finally, we are interested in reliability. Here you need to calculate an estimate of reliability of the test. Use (and reference) [@RN130]. Try to be clear with what measure of reliability you are using and what it is telling you.

## The format of the report

The report should be uploaded to github with both a raw .Rmd file and a .html file as output. The github folder should also contain the datset being used in the calculations. Contributions from members of the group can be made directly to the github directory. Follow this checklist on how to get started:

([A nice introduction to github with R can be found here](https://happygitwithr.com/install-intro.html))

- Create a [github account](http://www.github.com). All members of the group should have their own account.
- Once signed in to github.com, create a new repository with an informative name. Make sure that the repository is public. One member of the group can create this repository in their own account.
- Be sure to [download and install git](https://git-scm.com/downloads) on your computer. 
- Start up RStudio, start a new project, select the Version Control option and copy the address to the repository.
- Add files to your project and upload to github. You need to add files, commit them and push using git. Use the command line in your terminal:

```{r, eval = FALSE}
git add -A

git commit -m "A short message to describe your changes"

git push
```

- When you want to download the latest version of your project, type in the terminal:

```{r, eval = FALSE}

git pull

```

- You may encounter conflicts if pushing changes that overwrites changes made by other group members. These may be tricky but can be resolved... Have patience!

## How to hand in the report.

Copy the link to your github folder into canvas. 



