---
title: "IDR4000: Portfolio exam"
output: pdf_document
bibliography: resources/bib-final.bib
---

# Overall requirements for the portfolio exam

Your report should be a completely reproducible document, built from a source file (e.g. R markdown) with data and code stored in a repository. A reference/link to your github repository should be provided in the report. To make your life easier, I suggest that you compile your report from one repository. 



# Part 1: Reliability and tools for reproducible data science 

The purpose of this assignment is to present estimates of reliability of measures collected in the physiology lab. A second purpose is to use tools for reproducible data science. The report is a group collaboration.

## Elements of the report

The report should describe one test that you have performed in the physiology-lab. Select the test that is most interesting to you. The test should be described with a detailed protocol, including preparations of the participant (that is being tested), standardization of the test, and post-test data preparation. Post-test data preparation refers to steps needed to get data from e.g. equipment used during the test to your analysis. This section should take into account and reference [@RN1739; @RN2511].

The next section should contain descriptive data from the test. This could be measures of central tendency and variability in the measures you have collected. If possible, try to find similar estimates in the scientific literature.

Finally, we are interested in reliability. Here you need to calculate an estimate of reliability of the test. Use (and reference) [@RN130]. Try to be clear with what measure of reliability you are using and what it is telling you.

# Part 2: Laboratory report from DNA, RNA or protein extraction/analysis labs

This is a group collaboration. Each group is expected to present background, methods (including principles for measurements) and results from one of the labs (DNA, RNA or protein). You are free to shape the report as you see fit. 

# Part 3: Philosophy of science

This is an individual report. The report should be updated based on comments from Finnur. See a detailed description on Canvas.

# Part 4: Study designs

Choose an area of interest (e.g. protein supplementation for muscle hypertrophy or the effect of block
periodization on VO<sub>2max</sub>). Find at least five *original research studies*^[Avoid using review articles or meta-analyses] in your selected area and describe strength and weakness of these studies. The report should focus on the design of the studies and selection of statistical tests to answer study aims.

When analyzing your studies you can use the QALMRI method^[See [Course Notes](https://dhammarstrom.github.io/quant-methods/assignment4.html) and [Teaching undergraduate students to read empirical articles: An evaluation and revision of the QALMRI method](https://psyarxiv.com/p39sc/), this advice was also heavily influenced by [this website](https://bartlettje.github.io/2021-10-06-learning-to-read-scientific-journal-articles/)].

Your report should summarize results from the reviewed studies. Highlight differences and similarities between studies. *As the main points of this review are study designs and statistical analyses, this should be your main focus*. When doing your literature review, it is however a good idea to structure it in a table with the above mentioned headings:

- Question
- Alternative
- Logic
- Methods 
- Results
- Inference

Conclude your report with a recommendation, how should future studies in your area be designed to best
answer similar questions.


# Part 5: Analyzing repeated measures experiments {#assignment5}

In this assignment you will analyze and report on trial investigating the effect of resistance training volume on lean mass and muscle strength. The data are part of the `exscidata` package and can be accessed as `data("strengthvolume")` and `data("dxadata")`. Read the instructions carefully!

## Organizing the report

Your report should consist of the sections Introduction, Methods, Results and Discussion. Below follows detailed descriptions and requirements for each section.

### Introduction

This section should consist of a description of the field, resistance-training volume and muscle strength and mass. Use at least five to ten references to introduce your audience and explain why you are doing the analysis/study. A tip is to use the QALMRI method, introduced in Part 4, to structure the reading of background information. It is up to you how you motivate the study and how you phrase the purpose of the study. It could be a hypothesis based on previous studies, it could also be question to fill a knowledge gap that you have identified in your literature review.

Structure the introduction in paragraphs. A first paragraph could contain a general introduction to the field, why is it of interest to investigate resistance-training? A second paragraph could specifically describe the specific field of resistance-training volume, why is important to know more about how we are likely to respond to different training volumes. The second paragraph should incorporate definitions important for your report, e.g., training volume, muscle mass and strength. Try to incorporate these definition as a fluid part of the text.

A third (or last) paragraph of the introduction should contain a statement regarding the purpose of the study. The purpose could be descriptive, hypothesis-driven or guided by a question. Although it could be considered a bit backward, you should explore the data sets before you select your question/hypothesis/purpose for it to be possible to answer.

### Methods

The method should give a thorough overview of the study and specific details regarding data collection. You can read about the details of this specific study in [@RN2358]. Use your own words to describe the study based on this description. A nice way to structure the methods section is to include subheadings:

-   Participants and study overview: Describe the participants and give an overview of all tests/measurements. Participants should be described in the first table of the report (Table 1). The overview of the tests/measurements should be done without double presentation as details should be presented in subsequent sections.

-   Specific descriptions (e.g. strength tests): Describe in detail how tests/measurements that you mentioned in the overview where conducted.

-   Data analysis and statistics: Describe how you treated the data prior to statistical tests or procedures and what tests/procedures were used to draw inference (or more generally, to answer your purpose). Describe how you present data (e.g. descriptive data with mean (SD), inference with confidence intervals etc.).

### Results

Describe the results of your analysis. This description should make use of table and figures as well as a text that guides and structures the content to the reader. Think about it this way, the text should describe when and how to read the figures and tables. This means that all aspects of the results should be covered in the text. The figures and tables should also be "self explanatory", this means that you have to include descriptive figure captions and descriptions of tables (see below for tips).

As the main purpose of the analysis should concern the effect of training volume on muscle mass and strength, it is natural that the comparison of training outcomes between volume conditions is the main analysis in the results. You may also have questions regarding the relationship between muscle strength and mass gains, if there are differences between men and women etc. Selection of statistical/analysis techniques should reflect the study question/purpose.

### Discussion

Structure the discussion with a first paragraph describing the main results of the analysis, this could be the answer to your question or a statement regarding the study hypothesis. In the following paragraphs discuss all results that you have presented in the light of previous studies. It is your job to give the reader plausible interpretations and explanations of your results. This is how single scientific results are incorporated in our collective understanding. These interpretations can later be challenged, however if you give the reader good arguments and clear descriptions, your insights will be valuable to collective reasoning even if they turn out to be wrong in light of new data.

End the discussion with a summary or conclusion. Some journals request that you state your conclusions under a specific heading in the end of the report/article.

## Organizing the data analysis

### Data preparation

The data is already structured in the `exscidata` package. To access the data, use the following code:

```{r, eval = TRUE, echo = TRUE}

library(exscidata)
data("dxadata"); data("strengthvolume")


```

To get and overview of the variables in each data set use `?strengthvolume` and `?dxadata`. In the `dxadata` the variables of interest are organized in a more convenient way using the code below:

```{r, message = FALSE, warning = FALSE, echo = TRUE}

library(tidyverse)

dxadata %>%
  select(participant:include, lean.left_leg, lean.right_leg) %>%
  pivot_longer(names_to = "leg", 
               values_to = "lean.mass", 
               cols = lean.left_leg:lean.right_leg) %>%
  mutate(leg = if_else(leg == "lean.left_leg", "L", "R"), 
         sets = if_else(multiple == leg, "multiple", "single")) %>%
  select(participant, time, sex, include, sets, leg, lean.mass) %>%
  print()

```

# References



